Simulating High-Dimensional Multivariate Data using the \pkg{bigsimr} Package

### 2021-06-02 010-background.Rmd
get bookdown cross refs working

```{r tab:ch010-realDataTab}
set.seed(2020-02-25)
num_genes    <- 3
num_patients <- 5
gene_sample  <- sample(example_genes[1:1000], num_genes)

@@

To place `bigsimr`'s implementation and algorithms in context with existing approaches, here we qualitatively compare to other `R` multivariate simulation packages.  The `MASS` packages `mvnorm()` has long been available to produce random vectors from multivariate normal (and $t$) distributions. This familiar procedure cannot simulate from arbitrary margins and dependency measures, as well as fails to readily scale to high dimensions. Recently, the high-performance package `mvnfast` [@Fasiolo2016] has been released that provides HD multivariate normal generation, but this lacks flexibility in marginal models and dependency modeling. For dependent discrete data, LD packages exist, such as `GenOrd`. For copula computation, the full-featured package `copula` provides many different copulas and a fairly general LD random generator `rCopula()`. High-dimesional random vectors via `copula` takes an impractical amount of computing time [@Li2019gpu]. The `nortaRA` R package matches Pearson coefficients near exactly and is reasonable for LD settings. Finally, we know of no other `R` package that can produce HD random vectors with flexible margins and dependence modeling. Table (\#tab:compare-table) summarizes the above discussion.

\begin{table}[]
\centering
\caption{Selected feature comparison of bigsimr to existing R packages for multivariate simulation.}
\label{tab:compare-table}
\begin{tabular}{@{}llll@{}}
R Package & Scales to HD & Hetereogenous Margins & Flexible Dependence \\ \midrule
MASS      & No           & No                    & No                  \\
mvnfast   & Yes          & No                    & No                  \\
GenOrd    & No           & Discrete margins      & Yes                 \\
copula    & No           & Yes                   & No                  \\
nortaRA   & No           & Yes                   & No                  \\
bigsimr   & Yes          & Yes                   & Yes                
\end{tabular}
\end{table}

>>>>

```{r ch010-realDataTab}
set.seed(2020-02-25)
num_genes    <- 3
num_patients <- 5
gene_sample  <- sample(example_genes[1:1000], num_genes)

@@

To place `bigsimr`'s implementation and algorithms in context with existing approaches, here we qualitatively compare to other `R` multivariate simulation packages.  The `MASS` packages `mvnorm()` has long been available to produce random vectors from multivariate normal (and $t$) distributions. This familiar procedure cannot simulate from arbitrary margins and dependency measures, as well as fails to readily scale to high dimensions. Recently, the high-performance package `mvnfast` [@Fasiolo2016] has been released that provides HD multivariate normal generation, but this lacks flexibility in marginal models and dependency modeling. For dependent discrete data, LD packages exist, such as `GenOrd`. For copula computation, the full-featured package `copula` provides many different copulas and a fairly general LD random generator `rCopula()`. High-dimesional random vectors via `copula` takes an impractical amount of computing time [@Li2019gpu]. The `nortaRA` R package matches Pearson coefficients near exactly and is reasonable for LD settings. Finally, we know of no other `R` package that can produce HD random vectors with flexible margins and dependence modeling. Table \@ref(tab:compare-table) summarizes the above discussion.

\begin{table}[]
\centering
\caption{Selected feature comparison of bigsimr to existing R packages for multivariate simulation.}
\label{compare-table}
\begin{tabular}{@{}llll@{}}
R Package & Scales to HD & Hetereogenous Margins & Flexible Dependence \\ \midrule
MASS      & No           & No                    & No                  \\
mvnfast   & Yes          & No                    & No                  \\
GenOrd    & No           & Discrete margins      & Yes                 \\
copula    & No           & Yes                   & No                  \\
nortaRA   & No           & Yes                   & No                  \\
bigsimr   & Yes          & Yes                   & Yes                
\end{tabular}
\end{table}

### 2021-06-02 040-monte-carlo-evaluation.Rmd
get bookdown cross refs working

<!-- 
Table: (\#tab:sims) Identical margin, bivariate simulation configurations to evaluate matching accuracy.
\begin{table}[]
\toprule
Simulation Reps $B$ & Correlation Types & Identical-margin 2D distribution \\ \midrule
$1,000$ & Pearson ($\rho_P$) & ${\bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ \\
$10,000$ & Spearman ($\rho_S$) & ${\bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ \\
$100,000$ & Kendall ($\tau$) & ${\bf Y} \sim MVNB(p = 3\times10^{-4}, r = 4,\rho_i), i=1,\ldots,100$ \\ \bottomrule
\end{tabular}
\end{table}
-->

>>>>

\begin{table}[]
\centering
\caption{Selected feature comparison of `bigsimr` to other existing `R` multivariate simulation packages.}
\label{compare-table}
\begin{tabular}{llll}
R Package & Scales to HD & Hetereogenous Margins & Flexible Dependence \\ \hline
MASS      & No           & No                    & No                  \\
mvnfast   & Yes          & No                    & No                  \\
GenOrd    & No           & Discrete margins      & Yes                 \\
copula    & No           & Yes                   & No                  \\
nortaRA   & No           & Yes                   & No                  \\
bigsimr   & Yes          & Yes                   & Yes                
\end{tabular}
\end{table}

### 2021-06-02 050-examples.Rmd
get bookdown cross refs working

```{r ch050-densityEvaluationECHO, echo=TRUE}
<<ch050-densityEvaluation>>
pHat
```

@@

```{r ch050-quadlossECHO, echo=TRUE, eval=FALSE}
# Simulate random vectors equal to the sample size
N <- nrow(example_brca)
# create m random vectors and estimate correlation
simRho <- replicate(n = m, expr = {
  tmpSim <- bs$rvec(N , R_X, nb_margins)
  bs$cor(tmpSim, bs$Spearman)
}, simplify = FALSE)
frobenius_loss <- sapply(simRho, function(R) norm( R - R_S, type = "F")  )
```

The `R` summary function supplies the mean-augmented five-number summary of the Frobenius loss distribution computed above.

```{r ch050-loss-summary, echo=TRUE}
summary(frobenius_loss)
```

This distribution could be compared to bespoke HD covariance estimators for guidance on whether the additional complexity and computation time are warranted.

>>>>

```{r ch050-densityEvaluationECHO, echo=TRUE}
pHat
```

@@

```{r ch050-quadlossECHO, echo=TRUE, eval=FALSE}
# Simulate random vectors equal to the sample size
N <- nrow(example_brca)
# create m random vectors and estimate correlation
simRho <- replicate(n = m, expr = {
  tmpSim <- bs$rvec(N , R_X, nb_margins)
  bs$cor(tmpSim, bs$Spearman)
}, simplify = FALSE)
```

The `R` summary function supplies the mean-augmented five-number summary of the Frobenius loss from the target. This distribution could be compared to bespoke HD covariance estimators for guidance on whether the additional complexity and computation time are warranted.

```{r ch050-loss-summary, echo=TRUE}
frobenius_loss <- sapply(simRho, function(R) norm( R - R_S, type = "F")  )
summary(frobenius_loss)
```

### 2021-06-02 article_bigsimr.tex
get bookdown cross refs working

\caption{\label{tab:tab:ch010-realDataTab}mRNA expression for three selected high-expressing genes, STAU1, FKBP1A, NME2, for the first five patients in the TCGA BRCA data set.}
\centering
\begin{tabular}[t]{lrrr}
\toprule
Patient ID & STAU1 & FKBP1A & NME2\\
\midrule
TCGA-A1-A0SB & 10440 & 11354 & 17655\\
TCGA-A1-A0SD & 21523 & 20221 & 14653\\
TCGA-A1-A0SE & 21733 & 22937 & 35251\\
TCGA-A1-A0SF & 11866 & 19650 & 16551\\
TCGA-A1-A0SG & 12486 & 12089 & 10434\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> bs$cor_bounds(margins[1], margins[2], bs$Pearson, n_samples = 1e6)
\end{CodeInput}
\begin{CodeOutput}
Julia Object of type NamedTuple{(:lower, :upper),Tuple{Float64,Float64}}.
(lower = -0.8818561157816974, upper = 0.8818503162627974)
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> d <- ncol(sim_nbinom)
R> B <- nrow(sim_nbinom)
R> y0 <- 1
R> pHat <- mean(apply(sim_nbinom, 1, function(Y) {
+   all(Y >= y0)
+ }))
R> pHat
\end{CodeInput}
\begin{CodeOutput}
[1] 0.6015
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # Simulate random vectors equal to the sample size
R> N <- nrow(example_brca)
R> # create m random vectors and estimate correlation
R> simRho <- replicate(n = m, expr = {
+   tmpSim <- bs$rvec(N , R_X, nb_margins)
+   bs$cor(tmpSim, bs$Spearman)
+ }, simplify = FALSE)
R> frobenius_loss <- sapply(simRho, function(R) norm( R - R_S, type = "F")  )
\end{CodeInput}
\end{CodeChunk}

The \texttt{R} summary function supplies the mean-augmented five-number
summary of the Frobenius loss distribution computed above.

\begin{CodeChunk}
\begin{CodeInput}
R> summary(frobenius_loss)
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
    675     813     974     998    1080    1851 
\end{CodeOutput}
\end{CodeChunk}

This distribution could be compared to bespoke HD covariance estimators
for guidance on whether the additional complexity and computation time
are warranted.

>>>>

\caption{\label{tab:ch010-realDataTab}mRNA expression for three selected high-expressing genes, STAU1, FKBP1A, NME2, for the first five patients in the TCGA BRCA data set.}
\centering
\begin{tabular}[t]{lrrr}
\toprule
Patient ID & STAU1 & FKBP1A & NME2\\
\midrule
TCGA-A1-A0SB & 10440 & 11354 & 17655\\
TCGA-A1-A0SD & 21523 & 20221 & 14653\\
TCGA-A1-A0SE & 21733 & 22937 & 35251\\
TCGA-A1-A0SF & 11866 & 19650 & 16551\\
TCGA-A1-A0SG & 12486 & 12089 & 10434\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> bs$cor_bounds(margins[1], margins[2], bs$Pearson, n_samples = 1e6)
\end{CodeInput}
\begin{CodeOutput}
Julia Object of type NamedTuple{(:lower, :upper),Tuple{Float64,Float64}}.
(lower = -0.8807818345178882, upper = 0.8817164736117079)
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pHat
\end{CodeInput}
\begin{CodeOutput}
[1] 0.6015
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # Simulate random vectors equal to the sample size
R> N <- nrow(example_brca)
R> # create m random vectors and estimate correlation
R> simRho <- replicate(n = m, expr = {
+   tmpSim <- bs$rvec(N , R_X, nb_margins)
+   bs$cor(tmpSim, bs$Spearman)
+ }, simplify = FALSE)
\end{CodeInput}
\end{CodeChunk}

The \texttt{R} summary function supplies the mean-augmented five-number
summary of the Frobenius loss from the target. This distribution could
be compared to bespoke HD covariance estimators for guidance on whether
the additional complexity and computation time are warranted.

\begin{CodeChunk}
\begin{CodeInput}
R> frobenius_loss <- sapply(simRho, function(R) norm( R - R_S, type = "F")  )
R> summary(frobenius_loss)
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
   26.0    28.5    31.2    31.4    32.9    43.0 
\end{CodeOutput}
\end{CodeChunk}
